---
title: "Practical 3 Answers"
subtitle: "Working with Data: Exploratory Data Analysis"
author: "Your Name Here"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    link-external-newwindow: true
    link-external-icon: true
editor: source
---

<!--
Before you start:

- This practical uses the `penguins` dataset from the `palmerpenguins`
  package — there's no .csv to download this week, just an install
  (see the README for instructions).
- See the README.md in this repository for package installation instructions
  and Git/GitHub tips.
- Render this document with the "Render" button in RStudio, or by running
  quarto::quarto_render("practical3-answers.qmd") in the R console.
-->

## Getting started

```{r}
#| label: setup
#| message: false
#| warning: false

# Load the packages you need for this practical.
# See the README for installation instructions if any of these are missing.
library(ggplot2)
library(dplyr)
library(tidyr)
library(naniar)
library(palmerpenguins)
library(GGally)
```

## Palmer Penguins

### Question 1

Use the `naniar` package to inspect missingness. Are there any variables where missingness is high and this should be monitored?

_You can use `naniar::gg_miss_upset(...)` or other functions you've come across but you should carefully interpret the output, and review the raw data to confirm.^[There may be some warnings that you can safely ignore.]_

```{r}
#| label: inspect-missingness

```




### Question 2

Summarise the data variables in the dataset. What variables exist? What are their ranges, means etc? Hint: you can start by using `glimpse()` and `summary()`. You may also look at the whole dataset using `View()` in RStudio.

```{r}
#| label: summarise-variables

```




## Basic Plotting

### Question 3

Using `geom_histogram()`, plot the distribution of `flipper_length_mm`. What shape does the distribution appear to have? Use the starter code below and change the theme with `theme_bw()`.

```{r}
#| label: histogram-flipper-length

ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_histogram()
```




### Question 4

The default bin width is rarely ideal. Experiment with the `binwidth` argument in `geom_histogram(...)`. How does your choice of bin width change your interpretation of the distribution?

```{r}
#| label: histogram-binwidth

```




### Question 5

Using `geom_bar()`, visualise the count of penguins belonging to each species. Which species is most common in this dataset?

```{r}
#| label: bar-species-count

# ggplot(...) + geom_bar(aes(x = ... ))

```




### Question 6

Modify your bar chart to map the `fill` aesthetic to `island`. What does the resulting stacked bar tell you about where each species was sampled? Try `position = "dodge"` in `geom_bar()`, which version is appropriate and when?

```{r}
#| label: bar-species-island

```




### Question 7

Use `geom_boxplot()` to compare the distribution of `body_mass_g` across the three species. Map `species` to the x-axis and `body_mass_g` to the y-axis.

```{r}
#| label: boxplot-mass-species

```

### Question 8

Map the `fill` aesthetic in your previous boxplot to `sex`. What happens to the number of boxes, and what new comparison does this enable?

```{r}
#| label: boxplot-mass-species-sex

```




### Question 9

Remove the missing values in the previous plot by filtering. Is this good practice? When should missing values be removed?

```{r}
#| label: boxplot-remove-missing

# add filter for other column with missing data
penguins |> filter(is.na(sex))

# Or (use carefully) look at tidyr::drop_na() 

```




## Advanced Plotting

### Question 10

Recreate the above plot using `geom_point()`. Describe the overall relationship between these two variables and its dependence on species.

```{r}
#| label: scatter-mass-species

```




### Question 11

Change the previous plot so that each species is plotted separately using `facet_wrap()`. When would using facets be more appropriate than colours to distinguish categories? Try using the `scales = "free"` argument for `facet_wrap()` — when would this be helpful or inappropriate?

```{r}
#| label: facet-species

```




### Question 12

Use the `ggscatmat()` function from `GGally` to plot a scatter plot matrix of pairwise relations between continuous variables.

```{r}
#| label: scatmat-continuous

```

### Question 13

Add colouring by species to the `ggscatmat()` plot. From the plot, identify which two variables are likely to best distinguish each species. Are there any anomalies in the data?

```{r}
#| label: scatmat-species-colour

```



