---
title: "Practical 3 Answers"
subtitle: "Working with Data: Exploratory Data Analysis"
author: "Your Name Here"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    link-external-newwindow: true
    link-external-icon: true
editor: source
---

<!--
Before you start:

- This practical uses the `penguins` dataset from the `palmerpenguins`
  package — there's no .csv to download this week, just an install
  (see the README for instructions).
- See the README.md in this repository for package installation instructions
  and Git/GitHub tips.
- Render this document with the "Render" button in RStudio, or by running
  quarto::quarto_render("practical3-answers.qmd") in the R console.
-->

## Getting started

```{r}
#| label: setup
#| message: false
#| warning: false

# Load the packages you need for this practical.
# See the README for installation instructions if any of these are missing.
library(ggplot2)
library(dplyr)
library(tidyr)
library(naniar)
library(palmerpenguins)
library(GGally)
```

## Palmer Penguins

### Question 1

Use the `naniar` package to inspect missingness. Are there any variables where missingness is high and this should be monitored?

_You can use `naniar::gg_miss_upset(...)` or other functions you've come across but you should carefully interpret the output, and review the raw data to confirm.^[There may be some warnings that you can safely ignore.]_

```{r}
#| label: inspect-missingness

# How many missing values for each variable
naniar::miss_var_summary(penguins)

# Where are the missing values in the data set
naniar::vis_miss(penguins)

# Which variables tend to be missing together
naniar::gg_miss_upset(penguins)


```




### Question 2

Summarise the data variables in the dataset. What variables exist? What are their ranges, means etc? Hint: you can start by using `glimpse()` and `summary()`. You may also look at the whole dataset using `View()` in RStudio.

```{r}
#| label: summarise-variables

glimpse(penguins)

summary(penguins)

penguins |> count(species)

penguins |> count(island)

penguins |> count(sex)

penguins |> count(year)


```




## Basic Plotting

### Question 3

Using `geom_histogram()`, plot the distribution of `flipper_length_mm`. What shape does the distribution appear to have? Use the starter code below and change the theme with `theme_bw()`.

```{r}
#| label: histogram-flipper-length

ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_histogram() + 
  theme_bw()

## The distribution is not unimodal (e.g., like a bell curve). It looks to be bimodal, or even trimodal, with a cluster of shorter flippers (~180-195mm) and a separate cluster of longer flippers (~210-230mm), and a less distinct middle. This is a hint that flipper length differs systematically by some other variable in the data rather than representing one homogeneous population.
```




### Question 4

The default bin width is rarely ideal. Experiment with the `binwidth` argument in `geom_histogram(...)`. How does your choice of bin width change your interpretation of the distribution?

```{r}
#| label: histogram-binwidth

ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_histogram(binwidth = 10) +
  ggtitle("binwidth = 10") +
  theme_bw()

ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_histogram(binwidth = 2) +
  ggtitle("binwidth = 2") +
  theme_bw()

ggplot(penguins, aes(x = flipper_length_mm)) + 
  geom_histogram(binwidth = 0.5) + 
  ggtitle("binwidth = 0.5") + 
  theme_bw()

## Choice of bin width can hide or reveal structure in the data:
## Too wide a bin (e.g. binwidth = 10, meaning collect in 10mm bins) smooths right over the multiple clusters and can make the data look unimodal.
## A moderate bin width (e.g. binwidth = 2) is what reveals the multimodal shape clearly.
## Too narrow a bin (e.g. binwidth = 0.5) startsshowing gaps and noise from individual measurement values.
```




### Question 5

Using `geom_bar()`, visualise the count of penguins belonging to each species. Which species is most common in this dataset?

```{r}
#| label: bar-species-count

ggplot(penguins, aes(x = species)) + 
  geom_bar() + 
  scale_x_discrete("Species") + 
  scale_y_continuous("Count") + 
  theme

```




### Question 6

Modify your bar chart to map the `fill` aesthetic to `island`. What does the resulting stacked bar tell you about where each species was sampled? Try `position = "dodge"` in `geom_bar()`, which version is appropriate and when?

```{r}
#| label: bar-species-island

# Stacked (default position = "stack")
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar() +
  scale_x_discrete("Species") +
  scale_y_continuous("Count") +
  scale_fill_discrete("Island") +
  ggtitle("Stacked version") +
  theme_bw()

ggplot(penguins, aes(x = species, fill = island)) + 
  geom_bar(position = "dodge") + 
  scale_x_discrete("Species") + 
  scale_y_continuous("Count") + 
  scale_fill_discrete("Island") + 
  ggtitle("Dodged version") +
  theme_bw()

# Dodged version 2
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar(position = position_dodge(preserve = "single")) +
  scale_x_discrete("Species") +
  scale_y_continuous("Count") +
  scale_fill_discrete("Island") +
  ggtitle("Dodged 2.0") +
  theme_bw()


```




### Question 7

Use `geom_boxplot()` to compare the distribution of `body_mass_g` across the three species. Map `species` to the x-axis and `body_mass_g` to the y-axis.

```{r}
#| label: boxplot-mass-species


ggplot(penguins, aes(x = species, y = body_mass_g)) + 
  geom_boxplot() + 
  scale_x_discrete("Species") + 
  scale_y_continuous("Body Mass in grams") + 
  theme_bw()


```

### Question 8

Map the `fill` aesthetic in your previous boxplot to `sex`. What happens to the number of boxes, and what new comparison does this enable?

```{r}
#| label: boxplot-mass-species-sex

ggplot(penguins, aes(x = species, y = body_mass_g, fill = sex)) + 
  geom_boxplot() + 
  scale_x_discrete("Species") + 
  scale_y_continuous("Body Mass in grams") + 
  theme_bw()


```




### Question 9

Remove the missing values in the previous plot by filtering. Is this good practice? When should missing values be removed?

```{r}
#| label: boxplot-remove-missing

penguins |>
  filter(!is.na(sex)) |>
  ggplot(aes(x = species, y = body_mass_g, fill = sex)) +
  geom_boxplot() +
  scale_x_discrete("Species") +
  scale_y_continuous("Body mass (g)") +
  scale_fill_discrete("Sex") +
  theme_bw()

```




## Advanced Plotting

### Question 10

Recreate the above plot using `geom_point()`. Describe the overall relationship between these two variables and its dependence on species.

```{r}
#| label: scatter-mass-species

penguins |> 
  filter(!is.na(sex)) |> 
  ggplot(aes(x = species, y = body_mass_g, colour = sex)) + 
  geom_point(position = position_jitter(width = 0.2), alpha = 0.6) + 
  scale_x_discrete("species") + 
  scale_y_continuous("body mass in grams") + 
  scale_colour_discrete("Sex") +
  theme_bw()
```




### Question 11

Change the previous plot so that each species is plotted separately using `facet_wrap()`. When would using facets be more appropriate than colours to distinguish categories? Try using the `scales = "free"` argument for `facet_wrap()` — when would this be helpful or inappropriate?

```{r}
#| label: facet-species

penguins |>
  filter(!is.na(sex)) |>
  ggplot(aes(x = sex, y = body_mass_g, colour = sex)) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.6) +
  facet_wrap(~species) +
  scale_x_discrete("Sex") +
  scale_y_continuous("Body mass (g)") +
  scale_colour_discrete("Sex") +
  theme_bw()

# Same plot as before, but letting each facet's y-axis scale independently
penguins |>
  filter(!is.na(sex)) |>
  ggplot(aes(x = sex, y = body_mass_g, colour = sex)) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.6) +
  facet_wrap(~species, scales = "free") +
  scale_x_discrete("Sex") +
  scale_y_continuous("Body mass (g)") +
  scale_colour_discrete("Sex") +
  theme_bw()

```




### Question 12

Use the `ggscatmat()` function from `GGally` to plot a scatter plot matrix of pairwise relations between continuous variables.

```{r}
#| label: scatmat-continuous

GGally::ggscatmat(penguins, columns = c("bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g"))

```

### Question 13

Add colouring by species to the `ggscatmat()` plot. From the plot, identify which two variables are likely to best distinguish each species. Are there any anomalies in the data?

```{r}
#| label: scatmat-species-colour

GGally::ggscatmat(
  penguins,
  columns = c("bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g"),
  color = "species"
)
```



